Explaining thermal performance in fish: interplay of ecology and physiology drive growth and fitness response

### 2018-07-02 draft_model_only.Rnw

\title{Beyond conceptual eco-physiological hypotheses: insights from a general size and trait-based model of ectotherm thermal performance}
\author{Philipp Neubauer\\Dragonfly Data Science, Wellington, NZ  \and 
        Ken H. Andersen\\Centre for Ocean Life
        Danish Technical University, Lyngby, Denmark}

@@

Increasing temperature under climate change is thought to affect physiology of ectotherms through increases in metabolic demands, leading to changes in species performance with concomitant effects on species ecology. Although intuitively appealing, the driving mechanism behind thermal performance is contested: thermal performance (e.g., growth) appears correlated with metabolic scope (i.e., oxygen availability for activity) for a number of species, but a substantial number of datasets do not support oxygen limitation of long-term performance. Whether or not oxygen limitations via the metabolic scope, or a lack thereof, have major ecological consequences remains a highly contested question. Here, we propose a general size and trait-based model of energy and oxygen budgets to determine the relative influence of thermal sensitivity of metabolic rates, oxygen limitation, and environmental conditions on thermal performance. We show that oxygen limitation is not necessary to explain performance curves that are correlated with metabolic scope. Oxygen can drastically limit performance and fitness at temperature extremes, but changes in thermal performance are primarily driven by the interplay of changing metabolic rates and species ecology (i.e., food aquisition and mortality). We find that resource availibility strongly modulates growth and fitness repsonses to temperature, with individuals in low-resource environments showing increased growth sensitivity and stronger fitness gradients with temperature. Furthermore, our model reveals that fitness is often minimised a low temperatures. This trend opposes trends in growth and efficiency, suggesting a potential explanation for the paradox that species often occur at lower temperatures than their growth-optimum. We suggest that this model provides a mechanistic underpinning that can provide general and realistic predictions about temperature impacts on ectotherm performance and effects of climate change on species with different metabolic and ecological traits.

@@

There is currently no accepted universal concept or general mechanistic model that explains ectotherm thermal performance and ecological responses to temperature (\cite{lefevre_are_2016, jutfelt_oxygen-and_2018}). In fish, two main theories are debated as potential general therories describing temeprature influences on ectotherms. The theory of Oxygen and Capacity Limitation of Temperature (OCLTT; \cite{portner_oxygen-and_2010}),  based on the general idea of oxygen limitation for major ecological processes (growth, fitness, etc; see below), provides a conceptual framework rather than a mechanistic theoretical framework, and its major assumptions have been called into question on numerous occasions (\cite[e.g., ][]{lefevre_are_2016, jutfelt_oxygen-and_2018}). The gill-oxygen limitation theory (GOLT) aims to provide a physiological justification for the use of a temperature and oxygen scaled version of the von Bertalanffi growth model as a mechanistic model of thermal performance. The validity of gill limitation of oxygen supply, as well as the ultimate limitation of growth by oxygen in all conditions, is strongly debated (\cite{brander_overconfidence_2013,pauly_sound_2017,lefevre_modelling_2017,lefevre_models_2017}). In addition, the von Bertalanffi growth model only encapsulates a highly abstract view of organism bio-energetics and ecology, with no reference to ecological rates such as food aquisition, mortality and reproduction. As such, it can only represent a very limited set of aspects of species ecology, and cannot on its own account for changes in ecological rates that are often observed with temperature (\cite{englund_temperature_2011,rall_universal_2012}).

To our knowledge, no general theoretical framework exists to quantitatively explain and predict changes in fundamental ecological rates, such as observed changes of attack rates with temperature (\cite{englund_temperature_2011,rall_universal_2012}), from fundamental physiological processes. Instead, and as advocated by the metabolic theory of ecology (\cite{brown_toward_2004}), ecological theory often treats ecological rates as being directly temperature dependant, without a direct link to the underlying physiological drivers (\cite[e.g., ][]{vucic-pestic_warming_2011, guiet_effects_2016}). However, although ecological rates seem to follow some general patterns in the response to temprature (\cite{englund_temperature_2011,rall_universal_2012}), there is also significant heterogeneity between species and trait groups. Since the primary effect of temperature of organisms is on individual physiology, a general model to explain ecological response should be grounded in physiology, and explicitly link physiological constraints and ecology.

@@

Our model assumes that physiology is described by two key budgets: the energy and oxygen budgets (\cite{holt_climate_2015,holt_climate_2014}). At a high level, we assume that animals will adapt activity levels to optimise available energy for growth and reproduction relative to mortality risk. Available energy is limited either by food capture, food processing capacity, or by available oxygen. We further assume that temperature acts directly on rates that are determined by enzymatic activity: digestive activity (via maximum consumption) and metabolic costs. Consequently, temperature only acts on ecological rates (e.g., actual feeding rates) via optimisation of activity levels.

\subsection{Model description}

The quantitative size- and trait-based thermal impact model lets ectotherms adjust the relative amounts of time ($\tau$) spent on metabolically costly activity and resting/hiding to optimise the net energy $P$ gain relative to natural mortality. Since both energy gain and loss are sensitive to temperature and oxygen limitations, both the activity level and the net energy gain will be subject to these environmental constraints. Their interplay thus determines available energy for growth and reproduction.

The model is written in terms of weight $w$ and oxygen balance equations at temperature $T$:

\begin{align}
f(w,T) &= \frac{\tau }{\tau  + \frac{h c_T w^{q-p}}{\gamma\Theta} } \label{eq:f} \\
P(w,T) &= S(w,T) - D(w,T) \\
  &=(1-\beta-\phi)f(w,T) h c_T w^q  - c_T k w^n - \tau \delta c_T k w  \\
P_{0_2}(w,T) &= S_{O_2}(w,T) - D_{O_2}(w,T) \\
        &= f_{O_2}(T)w^n - \omega \left( \beta f(w,T) h c_T w^q + c_T k w^n + \tau \delta c_T k w \right)
\end{align}

where $f(w,T)$ is the feeding level ([0,1]) at weight $w$ and temperature $T$, determined by consumption rate $\gamma w^p \Theta$ (search rate $\gamma w^p$ times prey availability $\Theta$), maximum consumption $h w^q$, and $\tau$, which can be interpreted as the proportion of time spent foraging (or proportion of maximum attack rate) or a multiplyer on a basal foraging speed, such that at $\tau=1$, foraging occurs at maximum speed. In the following, we will refer to $\tau$ as the activity fraction for sake of generality. Maximum consumption, determined by digestive (enzymatic) processes (\cite{jeschke_predator_2002, sentis_parsing_2013}), is assumed to scale with temperature according to the Arrhenius scaling $c_T = e^{E_a(T-T_0)/k \times T \times T_0}$. Available energy $P(w,T)$ is determined by supply ($S(w,T)$) from prey consumption ($f(w,T) h c_T w^q$), with $\beta$ a loss due to specific dynamic action (SDA, or heat increment; the energy spent absorbing food), and $\phi$ is the fraction of food excreted and egested. Metabolic demands ($D(w,T)$) are those of standard metabolism ($k w^n$), as well as active metabolism, scaled in units of standard metabolism as $\delta k w$, with $\delta$ determining the maginitude of activity cost relative to standard metabolic costs. The activity fraction $\tau$ determines the proportion of total metabolic costs that are due to active metabolism. Note that we assume that active metabolism scales directly with $w$, owing to muscular demands scaling approximately isometrically with weight (\cite{glazier_activity_2009,brett1965relation}). In practice, the scaling may not be strictly isometric owing to reduced drag at larger size (\cite[e.g., ][]{ware_bioenergetics_1978}), as well as non-strictly isometric growth of muscle mass and other relevant factors with weight (\cite{killen_little_2007}). However, the precise exponents of standard and foraging metabolism only alter predictions quantitatively, but do not change the over-all results.

The oxygen budget $P_{0_2}(w,T)$, determines physiological limits to aerobic activity. Demand ($D_{O_2}(w,T)$) is the sum of oxygen used for SDA (i.e., the conversion of food into energy, growth and reproduction - $\beta f h c_T w^q$) and ``active metabolism", the sum of standard (or resting) metabolism ($c_T k w^n$) and the metabolic costs of foraging ($\tau \delta c_T k w$), with $\omega$ determining amount of oxygen required per unit of metabolised feed. Metabolic scope is the difference between $S_{O_2}(w,T)=f_{O_2}(T)w^n$, the maximum amount of oxygen supplied at temperature $T$, and oxygen demand due to standard metabolism ($\omega c_T k w^n$). We assume that the maximum oxygen supply (MOS), follows a flexible dome-shaped function that can emulate both a dome-shaped MOS as well as MOS that increas continously up to a lethal temeprature (Figure \ref{fig:O2_fig}, Appendix A).

We assumed that fish will adjust their activity level to maximise $F$: maximize available energy relative to mortality: $\tau_{\text{opt}}$ = $\text{argmax}_{\tau_{w,T}}F(\tau_{w,T})$ = $\text{argmax}_{\tau_{w,T}}P(\tau_{w,T})/M(\tau_{w,T})$. This optimisation is also known as Gilliam's rule (\cite{gilliam_habitat_1987}) and represents a "short-sighted" fitness optimisation that does not account for future changes in conditions, and is appropriate for simulating opmisation in changing environments (\cite{sainmont_effective_2015}). We further assumed here that mortality scales with activity level and weight as $w^{q-1}$ (\cite{andersen_how_2009, hartvig_food_2011}), such that $M(w) = (\rho+\eta\tau) w^{q-1}$, where $ \rho$ is mortality at mass $w=1$ and $\tau=0$, that is, with no activity beyond that covered by standard metabolism. $M$ is then the effective mortality at realised levels of activity $\tau$ at weight $w$.

We assumed that the metabolic scope dictates the upper limit of this activity, such that at $\tau_{\text{max}}$, oxygen demand $D_{0_2}$ equals total supply $S_{0_2}$. Both temperature and oxygen will influence $\tau$, such that at a given temperature and oxygen concentration, $\tau_{w,T,O_2} = \text{min}\left( \tau_{\text{opt}},\tau_{\text{max}} \right)$, meaning we assumed that animals will adapt their effort to optimise fitness $F$ given temperature and oxygen constraints.

@@

Temperature affects growth via its effects on the energy budget and the investment of available (surplus) energy into reproduction and growth. The change in allocation to reproduction with size and age in variable environmental conditions is described by the maturation reaction norm, which is generally defined as the probability of maturing at a certain age under different growth conditions (\cite[e.g., ][]{dieckmann_probabilistic_2007}). We defined the reaction norm as the mid-point of a logistic allocation function that determines investment in reproduction as a function of age and size. The allocation was parametrised as 
$$\phi(z) = 1/(1+\exp(-(c*z))) \text{, where }$$
$$z(t, w_t) = (w_t-w^*)\times \text{cos}(\text{atan}(b))-t\times \text{sin}(\text{atan}(b))$$  
rotates the coordinate system about the slope $b$ of the reaction norm, $t$ is the age, $w$ is the mass, $w^*$ is the intercept of the reaction norm, and $c$ determines how rapidly energy allocation shifts from somatic growth to reproduction.

>>>>

\title{Explaining thermal performance in fish: interplay of ecology and physiology drive growth and fitness response}
\author{Philipp Neubauer\\Dragonfly Data Science, Wellington, NZ  \and 
        Ken H. Andersen\\Centre for Ocean Life
        Danish Technical University, Lyngby, Denmark}

@@

Increasing temperatures under climate change are thought to affect individual physiology of ectotherms through increases in metabolic demands, leading to changes in species performance with concomitant effects on species ecology. Although intuitively appealing, the driving mechanisms behind thermal performance is contested: thermal performance (e.g., growth) appears correlated with metabolic scope (i.e., oxygen availability for activity) for a number of species, but a substantial number of datasets do not support oxygen limitation of long-term performance. Whether or not oxygen limitations via the metabolic scope, or a lack thereof, have major ecological consequences remains a highly contested question. Here, we propose a general size and trait-based model of energy and oxygen budgets to determine the relative influence of thermal sensitivity of metabolic rates, oxygen limitation, and environmental conditions on thermal performance. We show that oxygen limitation is not necessary to explain performance curves that are correlated with metabolic scope. Oxygen can drastically limit performance and fitness at temperature extremes, but changes in thermal performance are primarily driven by the interplay of changing metabolic rates and species ecology. Furthermore, our model reveals that fitness trends with temperature likely oppose trends in growth and efficiency, suggesting a potential explanation for the paradox that species often occur at lower temperatures than their growth-optimum. We suggest that this general model provides a mechanistic underpinning that can provide general and realistic predictions about temperature impacts on ectotherm performance and effects of climate change on species with different metabolic and ecological traits.

@@

There are few general concepts and models that can explain thermal performance and ecological responses to temperature. To our knowledge, no general theoretical framework exists to quantitatively explain and predict changes in fundamental ecological rates, such as observed changes of attack rates with temperature (\cite{englund_temperature_2011,rall_universal_2012}), from fundamental physiological processes. Instead, and as advocated by the metabolic theory of ecology (\cite{brown_toward_2004}), ecological theory often treats ecological rates as being directly temperature dependant, without a direct link to the underlying physiological drivers (\cite[e.g., ][]{vucic-pestic_warming_2011, guiet_effects_2016}).

Although ecological rates seem to follow some general patterns in the response to temprature (\cite{englund_temperature_2011,rall_universal_2012}), there is also significant heterogeneity between species and trait groups. A phenomenological description that assumes a general ecological temperature response thus leads to difficulties with extrapolation when observed ecological responses are heterogeneous across species or other model components (\cite[e.g., fish in marine ecosystem models; ][]{guiet_effects_2016}). In this case, a deeper understanding of the underlying drivers of thermal responses may be necessary in order to derive general predictions about ecological responses to changing temperatures (\cite[e.g., ][]{vucic-pestic_warming_2011}). Since the primary effect of temperature of organisms is on individual physiology, a general model to explain ecological response should be grounded in physiology.

@@

Our model assumes that physiology is described by two key budgets: the energy and oxygen budgets (\cite{holt_climate_2015,holt_climate_2014}). We assume that animals will adapt activity levels to optimise available energy for growth and reproduction relative to mortality risk. Available energy is determined either by food capture, food processing capacity, or by available oxygen. We further assume that temperature acts directly on rates that are determined by enzymatic activity: digestive activity (via maximum consumption) and metabolic costs. Consequently, temperature only acts on ecological rates (e.g., actual feeding rates) via optimisation of activity levels.

\subsection{Model description}
Ectotherms adjust the relative amounts of time ($\tau$) spent on metabolically costly activity and resting/hiding to optimise the net energy gain relative to mortality. In the following, we refer to $\tau$ as the activity fraction for sake of generality.   Since both energy gain and metabolic losses are sensitive to temperature and oxygen limitations, both the activity level and the net energy gain will be subject to these environmental constraints. Their interplay thus determines available energy for growth and reproduction.

Net energy gain $P$ (mass per time) is the difference between supply $S$ and metabolic demands $D$, each being functions of body weight $w$ and temperature $T$:
\begin{align}
P(w,T) &= S(w,T) - D(w,T) \\
          &=(1-\beta-\phi)f(w,T) h c(T) w^q  - c(T) k w^n - \tau \delta c(T) k w  \\
  \label{eq:P}
\end{align}
where the feeding level ($[0:1]$)is given by a Holling type II function response:
\begin{align}
  f(w,T) &= \frac{\tau \gamma\Theta w^{p} }{\tau \gamma\Theta w^{p} + h c(T) w^q} }, \label{eq:f} \\
\end{align}
where $h c(T) w^q$ is the maximum consumption rate. The corresponding function for oxygen is given in Appendix A.  The feeding level ([0,1]) is determined by the fraction of time spent foraging $\tau$, foraging rate $\gamma w^p \Theta$ (search rate $\gamma w^p$ times prey availability $\Theta$) and maximum consumption $h c(T) w^q$.  The supply is discounted by the loss due to specific dynamic action  $\beta$ (SDA, or heat increment; the energy spent absorbing food), and $\phi$ is the fraction of food excreted and egested.

Metabolic demands ($D(w,T)$) are standard metabolism ($\propto k w^n$), which scales with exponent $n<1$, and active metabolism $\propto \tau k_a w$, which scales proportional to mass owing to muscular demands scaling approximately isometrically with weight (\cite{glazier_activity_2009,brett1965relation}, and the activity fraction \emph{(Important note: you cannot interpret $\delta$ as the factorial aerobic scope (FAS) when the two metabolisms scale with different exponents!. I therefore propose to rename the constant to $k_a$)}.

Temperature scaling of all metabolic rates (standard and active metabolism and maximum consumption rate) is determined by digestive (enzymatic) processes (\cite{jeschke_predator_2002, sentis_parsing_2013}) as $c(T) = e^{E_a(T-T_0)/(kTT_0)}$.

The oxygen budget $P_{0_2}(w,T)$ (oxygen per time) follows a similar form to the mass budget:
\begin{align}
  P_{0_2}(w,T) &= S_{O_2}(w,T) - D_{O_2}(w,T) \\
        &= s_(T) w^n - \omega c(T) \left( \beta f(w,T) h w^q + k w^n + k_a w \right).
\end{align}
Demand ($D_{O_2}(w,T)$) is the sum of oxygen used for all metabolic processes in Eq.~\ref{eq:P} (except assimilation losses), with $\omega$ being amount of oxygen required per mass.  The oxygen supply scales with body weight as $w^n$ multiplied by a flexible dome-shaped function ($s(t)$) \emph{I've changed this away from $f$, because $f$ was used above for a function that is in the range 0:1, while this one has a unit} that can emulate both a dome-shaped MOS as well as MOS that increas continously up to a lethal temeprature (Figure \ref{fig:O2_fig}, Appendix A).

Fish will adjust their activity fraction $\tau$ to maximise fitness. We use Gilliam's rule as a fitness proxy \cite{gilliam_habitat_1987}:
\begin{equation}
  \tau^* = \argmax_{\tau} \{ \frac{P(w,T)}{M(w)} \}.
\end{equation}
This optimisation represents a "short-sighted" fitness optimisation that does not account for future changes in conditions, and is appropriate for optimisation in stable environments (\cite{sainmont_effective_2015}).  Mortality scales with activity fraction and weight as $w^{q-1}$ (\cite{andersen_how_2009, hartvig_food_2011}):
\begin{equation}
  \label{eq:M}
  M(w) = (\rho+\eta\tau) w^{q-1},
\end{equation}
$\rho$ is mortality at mass $w=1$ and $\tau=0$, that is, with no activity beyond that covered by standard metabolism.  Activity is limited by available oxygen, such that $P_{O_2}$ is not allowed to be negative \emph{Is that a correct interpretation?}.

%We assumed that the metabolic scope dictates the upper limit of this activity, such that at $\tau_{\text{max}}$, oxygen demand $D_{0_2}$ equals total supply $S_{0_2}$. Both temperature and oxygen will influence $\tau$, such that at a given temperature and oxygen concentration, $\tau_{w,T,O_2} = \text{min}\left( \tau_{\text{opt}},\tau_{\text{max}} \right)$, meaning we assumed that animals will adapt their effort to optimise fitness $F$ given temperature and oxygen constraints.

@@

Temperature affects growth via its effects on the energy budget and the investment of available (surplus) energy into reproduction and growth. The change in allocation to reproduction with size and age in variable environmental conditions is described by the maturation reaction norm, which is generally defined as the probability of maturing at a certain age under different growth conditions (\cite[e.g., ][]{dieckmann_probabilistic_2007}). We defined the reaction norm as the mid-point of a logistic allocation function that determines investment in reproduction as a function of age and size. The allocation was parametrised as 
$$\phi(z) = 1/(1+\exp(-(c z))) \text{, where }$$
$$z(t, w_t) = (w_t-w^*)\times \text{cos}(\text{atan}(b))-t\times \text{sin}(\text{atan}(b))$$  
rotates the coordinate system about the slope $b$ of the reaction norm, $t$ is the age, $w$ is the mass, $w^*$ is the intercept of the reaction norm, and $c$ determines how rapidly energy allocation shifts from somatic growth to reproduction.
